---
title: "Validate multiple expansions"
author: "David Helekal"
date: "12/2/2020"
output:
  html_document:
    df_print: paged
---

```{r, eval mcmc}
eval_mcmc <- F
```

```{bash, clean directory, include=FALSE, eval=eval_mcmc}
rm -rf ./Paper/Sim/sim_validate_multiple_exp
rm -rf ./Paper/Sim/tmp3

mkdir ./Paper/Sim/tmp3
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(CaveDive)
library(rmutil)
library(ape)
library(ggplot2)
library(ggtree)
library(ggExtra)
library(treeio)
library(viridis)
library(rjson)
library(egg)
library(scales)

dir.create(file.path("./Paper/Sim/", "sim_validate_multiple_exp"))
base_dir <- "./Paper/Sim/sim_validate_multiple_exp"
```
## Simulation Validation

This notebook simulates several different sets of genealogies containing a number of expansions and tests MCMC inference against them, aiming to validate whether the correct number of expansions can be recovered.

### Number of expansions recovery

This section validates how well the correct number of expansions can recovered. We consider genealogies with 1 to 5 expansions and 60 tips + 40 per expansion  

### Set up the experiment 
```{r, set parameters}
set.seed(1)

burn_in <- 0.1
n_it <- 2e7
thinning <- n_it/1e4  

repetitions <- 25
dims <- c(2:5)
lambdar <- 5
pop <- 5
```

```{r, generate parameter specification files, eval=eval_mcmc}
dir.create(file.path(base_dir, "test_spec"))
dir.create(file.path(base_dir, "test_out"))

seed <- 0

for(i in c(1:length(dims))) {
  dir_name <- paste0("dim_",i)
  dir.create(file.path(paste0(base_dir,"/test_spec"), dir_name))
    dir.create(file.path(paste0(base_dir,"/test_out"), dir_name))
    for (l in c(1:repetitions)) {
      params <- c(paste0("-e ", dims[i]),
                  paste0("-s ", seed),
                  paste0("-t ", 60+dims[i]*40), 
                  paste0("-l ", lambdar),
                  paste0("--meanscale ", pop),
                  paste0("--sdscale ", 1/2),
                  paste0("--nu ", 1/3),
                  paste0("--kappa ", 1/4),
                  paste0("--sdk ", 0.5),
                  paste0("-o ",base_dir,"/test_out/",dir_name, "/r_",l),
                  paste0("--metadata ", paste0("\"","dim_idx:",i,"\""))) 
      fname <- paste0("dim_",i,"_r_",l,".txt")
      fileConn<-file(paste0(base_dir,"/test_spec/", dir_name,"/", fname))
        writeLines(params, fileConn)
      close(fileConn)
      seed <- seed+1
  }
}
```

### Simulate the parameters of individual processes
```{bash, simulate parameters, results=FALSE, eval=eval_mcmc}
find ./Paper/Sim/sim_validate_multiple_exp | grep .*.txt | xargs -L 1 cat | parallel -j 10 --verbose --tmpdir ./Paper/Sim/tmp3 -n 11 --halt-on-error 2 eval Rscript "./Scripts/simulate_tree.R {}"
```

### Run MCMC inference 
```{bash, run mcmc, results=FALSE, eval=eval_mcmc}
find ./Paper/Sim/sim_validate_multiple_exp | grep .*tree.nwk | xargs dirname | parallel -j 10 --verbose --tmpdir ./Paper/Sim/tmp3 --halt-on-error 2 eval Rscript "./Scripts/run_expansions_inference_nwk.R -n 2e7 -t 2e3 --lambdar 5 -s 1 -f {}/tree.nwk -o {}"
```

### Produce plots for dimension mode analysis
```{r, process data}
experiment_specs <- list.files(path = paste0(base_dir,"/test_out"), pattern=".*tree_params.json", full.names = TRUE, recursive = TRUE)
experiment_dirs <- dirname(experiment_specs)

results_data <- lapply(c(1:length(experiment_specs)), function(i){
  sim_data <- fromJSON(file=experiment_specs[i])
  meta_data <- sim_data$meta
  sim_data <- lapply(sim_data, unlist) 
  sim_data <- lapply(sim_data, as.numeric) 

  dim.gt <- sim_data$n_exp
  ### Check that we got the number of expansions approximately right  
  expansions <- readRDS(paste0(experiment_dirs[i],"/expansions.rds"))
  expansions <- discard_burn_in(expansions, proportion=burn_in)
  pre <- expansions$phylo_preprocessed

  mcmc_data <- expansions$model_data
  event_data <- expansions$expansion_data

  p_correct_dim <- length(which(mcmc_data$dim==dim.gt))/length(mcmc_data$dim)
  expected_dim <- mean(mcmc_data$dim)

  unique.dim <- unique(mcmc_data$dim)
  mode_dim <- unique.dim[which.max(sapply(unique.dim, function(x) length(which(mcmc_data$dim == x))))]

  return(list(p_correct_dim=p_correct_dim, 
              expected_dim=expected_dim, 
              mode_dim=mode_dim, 
              dim_gt=dim.gt))
})
```

```{r, aggregate histograms}
experiment_specs <- list.files(path = paste0(base_dir,"/test_out"), pattern=".*tree_params.json", full.names = TRUE, recursive = TRUE)
experiment_dirs <- dirname(experiment_specs)

histogram_dfs <- lapply(c(1:length(experiment_specs)), function(i){
  sim_data <- fromJSON(file=experiment_specs[i])
  meta_data <- sim_data$meta
  sim_data <- lapply(sim_data, unlist) 
  sim_data <- lapply(sim_data, as.numeric) 

  dim.gt <- sim_data$n_exp
  ### Check that we got the number of expansions approximately right  
  expansions <- readRDS(paste0(experiment_dirs[i],"/expansions.rds"))
  expansions <- discard_burn_in(expansions, proportion=burn_in)

  mcmc_data <- expansions$model_data
  return(data.frame(dim=mcmc_data$dim, gt=dim.gt))
})
```

```{r, create dataframe}
names(results_data) <- c(1:length(results_data)) 
data_df <- do.call(rbind.data.frame, results_data)

hist_df <- do.call(rbind, histogram_dfs)
hist_df$gt <- factor(hist_df$gt, levels=c(2:5), ordered=T)
head(hist_df)
```

```{r, plot mode dim, eval=T, out.width="90%", fig.width=20, fig.height=20, dpi=300, fig.align="center"}

e_hist <- ggplot(hist_df, aes(y=dim)) + 
          geom_bar(aes(x = ..prop..), stat="count") + 
          scale_y_discrete(limits=c(0:9)) + 
          scale_x_continuous(labels=percent, limits=c(0,1))+
          coord_cartesian(ylim = c(-0.5, 9.5), xlim=c(0,0.5),expand = FALSE) +
          ylab("Number of Expansions") + 
          facet_grid(cols=vars(gt)) +
          theme_bw() +
          theme(text = element_text(size=35),  
                axis.title.x = element_blank(),
                axis.text.x = element_blank(), 
                axis.ticks.x = element_blank())

max_e <- max(data_df$expected_dim)

e_dim_scatter <- ggplot(data_df, aes(dim_gt, expected_dim))
e_dim_scatter <- e_dim_scatter + geom_boxplot(aes(group=dim_gt)) + stat_summary(fun.y=mean, geom="point", lwd=2, color="red")
e_dim_scatter <- e_dim_scatter + geom_jitter(shape=16, dotsize=1, position=position_jitter(0.2))
e_dim_scatter <- e_dim_scatter + theme_bw() + labs(x = "True Number of Expansions", y="Expected Number of Expansions") + scale_x_discrete(limits=c(2:5)) + scale_y_continuous(limits=c(0, 12), breaks=c(0:12)) + coord_cartesian(ylim = c(-0.5, 9.5),xlim=c(1.5,5.5), expand = FALSE)
e_dim_scatter <- e_dim_scatter + theme(axis.text.x = element_text(angle = 45, hjust = 1),
                                        text = element_text(size=35),  
                                        axis.title.y = element_blank(),
                                        axis.title.x = element_blank(), 
                                        axis.text.y = element_blank(), 
                                        axis.ticks.y = element_blank())


p <- ggarrange(e_hist, e_dim_scatter, widths=c(2,2), heights=c(2))
png("./Paper/Figures/fig4.png", width=1600,height=800)
p
dev.off()
```